An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2016-10-12 aistats2017.tex
Shortened paper a bit, main part is now 8.7 pages.

The acceptance test guarantees detailed balance, which means
$p(\theta)p(\theta'\mid\theta) = p(\theta')p(\theta \mid\theta')$, where
$p(\theta'\mid\theta)$ is the probability of a transition from state $\theta$ to
$\theta'$. Here, $p(\theta'\mid\theta)=q(\theta'\mid\theta)\alpha(\theta,\theta')$.
% Daniel: I commented this out because people should know what it means and we
%re-state another detailed balance equation later!
%\begin{equation}\label{eq:detailed_balance2}
%    p(\theta)q(\theta'\mid\theta)\alpha(\theta,\theta') = p(\theta')q(\theta\mid\theta')\alpha(\theta',\theta)
%\end{equation}
This condition, together with ergodicity, guarantees that the Markov chain has a
unique stationary distribution $\pi(\theta) = p(\theta)$. For Bayesian
inference, we would like to sample from a parameter distribution for $\theta$
based on some observed data, so the target distribution is $p(\theta \mid x_1,
\ldots, x_N)$. The acceptance probability is now:
\begin{equation}\label{eq:acceptance_probability}
    \alpha(\theta,\theta') = 
    \frac{p_0(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta \mid
    \theta')}{p_0(\theta)\prod_{i=1}^N p(x_i \mid \theta)q(\theta' \mid\theta)}
    \wedge 1
\end{equation}
where $p_0(\theta)$ is a prior, and $p(x_i \mid \theta)$ are the probabilities
of the observations. Computing samples this way requires the use of all $N$
training data points, but this is very expensive for large datasets. To address
this challenge,~\citet{cutting_mh_2014,icml2014c1_bardenet14} perform approximate
Metropolis-Hasting tests using sequential hypothesis testing. During each
iteration, they start with a small minibatch of data and test whether the sample $\theta'$ 
should be accepted based on an approximate version of
the test $u < \alpha(\theta,\theta')$. If the approximate test cannot make a
decision with sufficient confidence, then the minibatch size is increased and
the test repeats. This process continues until a decision. The bounds depend on
either an asymptotic Central Limit Theorem~\citep{cutting_mh_2014} or a
concentration bound~\citep{icml2014c1_bardenet14}. The latter requires direct
bounds on the log likelihood ratio, which for general distributions requires
knowing $p(x_i \mid \theta)$ and $p(x_i \mid \theta')$ for all $N$ samples. In
addition, both methods suffer the drawback of resolving small log likelihood
ratio differences between the minibatch and full batch versions.  In the worst
case, all $N$ data points may be needed, as we show in
Section~\ref{ssec:gaussian_example}.

@@

The problem is that these methods use tail bounds to separate $\Delta$ away
from zero, but for certain input/random $u$ combinations, $\Delta$ can be
arbitrarily close to zero. We will avoid this by using the {\em approximately
normal} variation in $\Delta^*$ to {\em replace} the variation due to $u$.

\subsection{MCMC Posterior Inference}
There is a separate line of MCMC work drawing principles from statistical
physics. By viewing random variables as particles in a system, one can apply
Hamiltonian Monte Carlo (HMC)~\citep{mcmc_hamiltonian_2010} methods which
generate high acceptance \emph{and} distant proposals when run on full batches
of data. Recently Langevin Dynamics~\citep{langevin_2011,conf/icml/AhnBW12} has
been applied to Bayesian estimation on minibatches of data. This simplified
dynamics uses local proposals and avoids M-H tests by using small proposal steps
whose acceptance approaches 1 in the limit. However, the constraint on proposal
step size is severe, and the state space exploration reduces to a random walk.
Full minibatch HMC for minibatches was described in~\citet{sghmc_2014} which
allows momentum-augmented proposals with larger step sizes. However, step sizes
are still limited by the need to run accurately without M-H tests.  By providing
an M-H test with similar cost to standard gradient steps, our work opens the
door to applying those methods with much more aggressive step sizes without loss
of accuracy.

@@

\section{COMPUTING THE CORRECTION DISTRIBUTION}\label{sec:correction}

Our proposed test in Equation~(\ref{eq:criteria}) requires knowing the
distribution of the correction variable $X_{\rm corr}$ such that $X_{\rm norm} +
X_{\rm corr} = X_{\rm log}$, where $X_{\rm norm} \sim \mathcal{N}(0,\sigma^2)$
and $X_{\rm log}$ has a standard logistic CDF, $(1+\exp(-X))^{-1}$. In
Section~\ref{sec:analysis}, we show that the accuracy of the test depends on the
absolute error between the CDFs of $X_{\rm norm} + X_{\rm corr}$ and $X_{\rm
log}$. Consequently, we need to minimize this in our construction of $X_{\rm
corr}$. More formally, let $\Phi_{s_X} = \Phi(X/s_X)$ where $\Phi$ is the
standard normal CDF\footnote{Hence, $\Phi_{s_X}$ is the CDF of a zero-mean
Gaussian with standard deviation $s_X$.}, $S(X)$ be the logistic function, and
$C_{\sigma}(X)$ be the \emph{density} of the correction $X_{\rm corr}$
distribution. Our goal, based on Equation~(\ref{eq:deconvolution}), is to solve
the following optimization problem:
\begin{equation}\label{eq:overall_corr_problem}
    C_\sigma^* = \argmin_{C_\sigma} |\Phi_{\sigma} * C_{\sigma} - S|
\end{equation}
where $*$ denotes convolution.

@@

In the most similar prior works,~\citet{cutting_mh_2014} uses asymptotic
arguments based on the CLT to argue that its approximate acceptance test error
tends to zero as batch size increases, but no quantitative bounds are given.
In~\citet{icml2014c1_bardenet14}, explicit bounds are given, but they depend on
bounding:
\begin{equation}\label{eq:bad_bound}
    C_{\theta, \theta'} = \max_{1\leq i\leq N}|\log p(x_i\mid\theta') - \log p(x_i\mid\theta)|.
\end{equation}
Such bounds can be derived efficiently for models such as logistic regression,
but it is unclear how to derive them for a complex model such as a neural
network. In general, since a new $\theta'$ value is obtained each iteration, one
would need to iterate through all the $p(x_i\mid \theta')$ terms\footnote{The
sample code provided by the authors of~\citet{icml2014c1_bardenet14} in fact
computes $C_{\theta,\theta'}$ explicitly for each sample generated, i.e. it
traverses the entire data, thus providing no performance advantage over the
complete test.}. In contrast, we use quantitative forms of the Central Limit
Theorem which rely on measurable statistics from a single minibatch. Thus a
sampler using our approach does not need to see data beyond the current
minibatch.

@@

\begin{lemma}\label{lem:cdf_bounds}
Let $P(x)$ and $Q(x)$ be two cumulative distributions satisfying
$\sup_x|P(x)-Q(x)|\leq \epsilon$ with $x$ in some real range. Let $R(y)$ be the
{\em density} of another random variable $y$. Let $P'$ be the convolution $P*R$
and $Q'$ be the convolution $Q*R$. Then $P'(z)$ (resp. $Q'(z)$) is the CDF of
sum $z=x+y$ of independent random variables $x$ with CDF $P(x)$ (resp. $Q(x)$)
and y with density $R(y)$.  Then
\begin{equation}
    \sup_x|P'(x)-Q'(x)|\leq \epsilon
\end{equation}
\end{lemma}

@@

We start with a simple Gaussian mixture model, borrowing an experiment
from~\citet{langevin_2011}.  The parameter is 2-D, $\theta =
\langle \theta_1,\theta_2 \rangle$, and the parameter/data generation process is
\begin{equation}\label{eq:data_generation}
\begin{split}
    \theta &\sim \mathcal{N}(0, {\rm diag}(\sigma_1^2,\sigma_2^2)) \\
    x_i & \sim 0.5 \cdot \mathcal{N}(\theta_1, \sigma_x^2) + 0.5 \cdot \mathcal{N}(\theta_1+\theta_2, \sigma_x^2).
\end{split}
\end{equation}
We set $\sigma_1^2 = 10, \sigma_2^2 = 1$ and $\sigma_x^2=2$.  We fix $\theta =
\langle 0,1 \rangle$. The original paper sampled 100 data points from this
distribution, and estimated the posterior in the parameters. We are interested
in performance on larger problems and so sampled 1,000,000 points to form the
posterior of $p(\theta)\prod_{i=1}^{1,000,000}p(x_i\mid \theta)$ with the same
prior from Equation~(\ref{eq:data_generation}). This produces a much sharper
posterior with two very narrow peaks.  Our goal is to reproduce the original
posterior, so we adjust the temperature to $T=10,000$.  Taking logs, we get the
target as shown in the far left of Figure~\ref{fig:gauss_mix_1}.

@@

For large $n$, the per-bin distributions are well-approximated by Poisson
distributions with parameter $\lambda_i=P_i n$. Given a set of samples
$\{\theta_1,\ldots,\theta_T\}$, let $c_j$ denote the number of individual
samples $\theta_i$ that fall in bin $j$ out of $N_{\rm bins}$ total. We have
\begin{equation}\label{eq:log_prob_poissons}
\begin{split}
    \log p(c_1, \ldots, c_{N_{\rm bins}} \mid P_1, \ldots, P_{N_{\rm bins}}) & =  \\
    \sum_{j=1}^{N_{\rm bins}} c_j \log (n P_j) - n P_j - \log(\Gamma(c_j+1)). & \\
\end{split}
\end{equation}
The results for each model are shown in Table~\ref{tab:poissons}.

While the Poisson results provide general guidance on the relative accuracy of
the models, it is difficult to interpret the scores. Thus, we perform
significance tests to show the difference between the MCMC-sampled distributions
and the ground-truth using the Chi-Squared distribution as the test statistic,
which we also show in Table~\ref{tab:poissons}. Both results imply that our
method is slightly superior to~\citet{cutting_mh_2014}, but slightly worse
than~\citet{icml2014c1_bardenet14}.

Figure~\ref{fig:gauss_mix_2}, however, suggests that our method dominates in
terms of speed and efficiency. It shows histograms of the (final) minibatch
sizes used each iteration by the three methods. Our method consumes
significantly less data; most minibatch sizes are smaller than 1000, and the
average size is 210.  The other methods occasionally need to consume nearly all
data points, and average minibatch sizes are 15562 and 16857. The average
minibatch sizes roughly predict the running times of these methods since all
have a running time proportional to the total data consumed, with the exception
of~\citet{icml2014c1_bardenet14}, which requires a pass over the data to compute
$C_{\theta,\theta'}$ (see Equation~(\ref{eq:bad_bound})). Using minibatch sizes
as a proxy for time also avoids discrepancies due to code optimization.

>>>>

The acceptance test guarantees detailed balance, which means
$p(\theta)p(\theta'\mid\theta) = p(\theta')p(\theta \mid\theta')$, where
$p(\theta'\mid\theta)$ is the probability of a transition from state $\theta$ to
$\theta'$. Here, $p(\theta'\mid\theta)=q(\theta'\mid\theta)\alpha(\theta,\theta')$.
% Daniel: I commented this out because people should know what it means and we
%re-state another detailed balance equation later!
%\begin{equation}\label{eq:detailed_balance2}
%    p(\theta)q(\theta'\mid\theta)\alpha(\theta,\theta') = p(\theta')q(\theta\mid\theta')\alpha(\theta',\theta)
%\end{equation}
This condition, together with ergodicity, guarantees that the Markov chain has a
unique stationary distribution $\pi(\theta) = p(\theta)$. For Bayesian
inference, the target distribution is $p(\theta \mid x_1, \ldots, x_N)$. The
acceptance probability is now:
\begin{equation}\label{eq:acceptance_probability}
    \alpha(\theta,\theta') = 
    \frac{p_0(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta \mid
    \theta')}{p_0(\theta)\prod_{i=1}^N p(x_i \mid \theta)q(\theta' \mid\theta)}
    \wedge 1
\end{equation}
where $p_0(\theta)$ is the prior. Computing samples this way requires the use of all $N$
training data points, but this is very expensive for large datasets. To address
this challenge,~\citet{cutting_mh_2014,icml2014c1_bardenet14} perform approximate
Metropolis-Hasting tests using sequential hypothesis testing. During each
iteration, they start with a small minibatch of data and test whether the sample $\theta'$ 
should be accepted based on an approximate version of
the test $u < \alpha(\theta,\theta')$. If the approximate test cannot make a
decision with sufficient confidence, then the minibatch size is increased and
the test repeats. This process continues until a decision. The bounds depend on
either an asymptotic Central Limit Theorem~\citep{cutting_mh_2014} or a
concentration bound~\citep{icml2014c1_bardenet14}. The latter requires direct
bounds on the log likelihood ratio, which for general distributions requires
knowing $p(x_i \mid \theta)$ and $p(x_i \mid \theta')$ for all $N$ samples. In
addition, both methods suffer the drawback of resolving small log likelihood
ratio differences between the minibatch and full batch versions.  We discuss
a worst-case scenario in Section~\ref{ssec:gaussian_example}.

@@

The issue is the use of tail bounds to separate $\Delta$ from zero; for certain
input/random $u$ combinations, $\Delta$ can be arbitrarily close to zero. We
avoid this by using the {\em approximately normal} variation in $\Delta^*$ to
{\em replace} the variation due to $u$.

\subsection{MCMC Posterior Inference}
There is a separate line of MCMC work drawing principles from statistical
physics. One can apply
Hamiltonian Monte Carlo (HMC)~\citep{mcmc_hamiltonian_2010} methods which
generate high acceptance \emph{and} distant proposals when run on full batches
of data. Recently Langevin Dynamics~\citep{langevin_2011,conf/icml/AhnBW12} has
been applied to Bayesian estimation on minibatches of data. This simplified
dynamics uses local proposals and avoids M-H tests by using small proposal steps
whose acceptance approaches 1 in the limit. However, the constraint on proposal
step size is severe, and the state space exploration reduces to a random walk.
Full minibatch HMC for minibatches was described in~\citet{sghmc_2014} which
allows momentum-augmented proposals with larger step sizes. However, step sizes
are still limited by the need to run accurately without M-H tests.  By providing
an M-H test with similar cost to standard gradient steps, our work opens the
door to applying those methods with much more aggressive step sizes without loss
of accuracy.

@@

% Daniel: Saves space by using one line instead of two. Lame, I know.
%\section{COMPUTING THE CORRECTION DISTRIBUTION}\label{sec:correction}
\section{CORRECTION DISTRIBUTION}\label{sec:correction}

Our test in Equation~(\ref{eq:criteria}) requires knowing the
distribution of $X_{\rm corr}$ such that $X_{\rm norm} +
X_{\rm corr} = X_{\rm log}$, where $X_{\rm norm} \sim \mathcal{N}(0,\sigma^2)$
and $X_{\rm log}$ has a standard logistic CDF, $(1+\exp(-X))^{-1}$. In
Section~\ref{sec:analysis}, we show that the test accuracy depends on the
absolute error between the CDFs of $X_{\rm norm} + X_{\rm corr}$ and $X_{\rm
log}$. Consequently, we need to minimize this in our construction of $X_{\rm
corr}$. More formally, let $\Phi_{s_X} = \Phi(X/s_X)$ where $\Phi$ is the
standard normal CDF\footnote{Hence, $\Phi_{s_X}$ is the CDF of a zero-mean
Gaussian with standard deviation $s_X$.}, $S(X)$ be the logistic function, and
$C_{\sigma}(X)$ be the \emph{density} of the correction $X_{\rm corr}$
distribution. Our goal is to solve:
\begin{equation}\label{eq:overall_corr_problem}
    C_\sigma^* = \argmin_{C_\sigma} |\Phi_{\sigma} * C_{\sigma} - S|
\end{equation}
where $*$ denotes convolution.

@@

In the most similar prior works,~\citet{cutting_mh_2014} uses asymptotic
arguments based on the CLT to argue that its approximate acceptance test error
tends to zero as batch size increases, but no quantitative bounds are given.
In~\citet{icml2014c1_bardenet14}, explicit bounds are given, but they depend on
bounding:
\begin{equation}\label{eq:bad_bound}
    C_{\theta, \theta'} = \max_{1\leq i\leq N}|\log p(x_i\mid\theta') - \log p(x_i\mid\theta)|.
\end{equation}
Such bounds can be derived efficiently for models such as logistic regression,
but it is unclear how to derive them for a complex model such as a neural
network. In general, since a new $\theta'$ value is obtained each iteration, one
would need to use all the $p(x_i\mid \theta')$ terms\footnote{The sample code
provided by the authors of~\citet{icml2014c1_bardenet14} in fact computes
$C_{\theta,\theta'}$ explicitly for each sample generated, i.e. it traverses the
entire data, thus providing no performance advantage over the complete test.}.
In contrast, we use quantitative forms of the Central Limit Theorem which rely
on measurable statistics from a single minibatch. Thus a sampler using our
approach does not need to see data beyond the current minibatch.

@@

\begin{lemma}\label{lem:cdf_bounds}
Let $P(x)$ and $Q(x)$ be two CDFs satisfying
$\sup_x|P(x)-Q(x)|\leq \epsilon$ with $x$ in some real range. Let $R(y)$ be the
{\em density} of another random variable $y$. Let $P'$ be the convolution $P*R$
and $Q'$ be the convolution $Q*R$. Then $P'(z)$ (resp. $Q'(z)$) is the CDF of
sum $z=x+y$ of independent random variables $x$ with CDF $P(x)$ (resp. $Q(x)$)
and y with density $R(y)$.  Then
\begin{equation}
    \sup_x|P'(x)-Q'(x)|\leq \epsilon
\end{equation}
\end{lemma}

@@

We borrow a Gaussian mixture experiment from~\citet{langevin_2011}.  The
parameter is 2-D, $\theta = \langle \theta_1,\theta_2 \rangle$, and the
generation process is
\begin{equation}\label{eq:data_generation}
\begin{split}
    \theta &\sim \mathcal{N}(0, {\rm diag}(\sigma_1^2,\sigma_2^2)) \\
    x_i & \sim 0.5 \cdot \mathcal{N}(\theta_1, \sigma_x^2) + 0.5 \cdot \mathcal{N}(\theta_1+\theta_2, \sigma_x^2).
\end{split}
\end{equation}
We set $\sigma_1^2 = 10, \sigma_2^2 = 1$ and $\sigma_x^2=2$.  We fix $\theta =
\langle 0,1 \rangle$. The original paper sampled 100 data points from this
distribution, and estimated the posterior in the parameters. We are interested
in performance on larger problems and so sampled 1,000,000 points to form the
posterior of $p(\theta)\prod_{i=1}^{1,000,000}p(x_i\mid \theta)$ with the same
prior from Equation~(\ref{eq:data_generation}). This produces a much sharper
posterior with two very narrow peaks.  Our goal is to reproduce the original
posterior, so we adjust the temperature to $T=10,000$.  Taking logs, we get the
target as shown in the far left of Figure~\ref{fig:gauss_mix_1}.

@@

For large $n$, the per-bin distributions are approximated by Poissons with
parameter $\lambda_i=P_i n$. Given samples $\{\theta_1,\ldots,\theta_T\}$, let
$c_j$ denote the number of individual samples $\theta_i$ that fall in bin $j$
out of $N_{\rm bins}$ total. We have
\begin{equation}\label{eq:log_prob_poissons}
\begin{split}
    \log p(c_1, \ldots, c_{N_{\rm bins}} \mid P_1, \ldots, P_{N_{\rm bins}}) & =  \\
    \sum_{j=1}^{N_{\rm bins}} c_j \log (n P_j) - n P_j - \log(\Gamma(c_j+1)). & \\
\end{split}
\end{equation}
The results for each model are shown in Table~\ref{tab:poissons}. It is
difficult, however, to interpret the scores. Thus, we perform significance tests
to show the difference between the MCMC-sampled distributions and the
ground-truth using the Chi-Squared distribution as the test statistic, which we
also show in Table~\ref{tab:poissons}. Both results imply that our method is
slightly superior to~\citet{cutting_mh_2014}, but slightly worse
than~\citet{icml2014c1_bardenet14}.

Figure~\ref{fig:gauss_mix_2}, however, suggests that our method dominates in
terms of speed and efficiency. It shows histograms of the (final) minibatch
sizes used each iteration. Our method consumes
significantly less data; most sizes are smaller than 1000, and the
average size is 210.  The other methods occasionally need to consume nearly all
data points, and average minibatch sizes are 15562 and 16857. The average
minibatch sizes roughly predict the running times of these methods since all
have a running time proportional to the total data consumed, with the exception
of~\citet{icml2014c1_bardenet14}, which requires a pass over the data to compute
$C_{\theta,\theta'}$ (see Equation~(\ref{eq:bad_bound})). Using minibatch sizes
as a proxy for time also avoids discrepancies due to code optimization.
